--- 
title: "DATA 101 Shortest Textbook"
subtitle: "How To Accomplish More With Less"
author: 
    name: "Tomasz Imielinski"
    email: "timielinski@gmail.com"
# subauthor: 
#     name: "Deep Lokhande"
#     email: "deep.lokhande@rutgers.edu"
# date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
# bibliography: [book.bib, packages.bib]
# biblio-style: apalike
# link-citations: yes
description: "This is a example of a interactive book for DATA 101 Course thought by Prof. Tomasz Imielinski (http://data101.cs.rutgers.edu/) Find the demo site at (https://deeplokhande.github.io/data101demobook/) Published using Bookdown for R."
---



# FRONTMATTER{-}

<script src=https://cdn.datacamp.com/datacamp-light-latest.min.js></script>
```{r ,include=FALSE}
tutorial::go_interactive( greedy = FALSE)

```


This is a textbook based on the [DATA 101](http://data101.cs.rutgers.edu/) course thought by Prof. Tomasz Imielinski at Computer Science Department at Rutgers University, New Brunswick.

Data 101 is an introductory course for beginners from any field of study, interested in the field of Data Science. 

Acknowledgment: We thank the significant contributions of [Deep Lokhande](deeplokhande.github.io) towards building this book.

Note: This book is undergoing constant updates following along with the course thought in Spring 2021. Topics under progress are marked with a " * ".








<!-- **FIRST FEW PAGES CAN BE SKIPPED. JUMP TO RECITAION 3 FOR DEMO** -->

<!-- This is a _sample_ book written in **Markdown**. You can use anything that Pandoc's Markdown supports, e.g., a math equation $a^2 + b^2 = c^2$. -->

<!-- The **bookdown** package can be installed from CRAN or Github: -->

<!-- ```{r eval=FALSE} -->
<!-- install.packages("bookdown") -->
<!-- # or the development version -->
<!-- # devtools::install_github("rstudio/bookdown") -->
<!-- ``` -->

<!-- Remember each Rmd file contains one and only one chapter, and a chapter is defined by the first-level heading `#`. -->

<!-- To compile this example to PDF, you need XeLaTeX. You are recommended to install TinyTeX (which includes XeLaTeX): <https://yihui.org/tinytex/>. -->

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```
